---
title: "Hands-on Exercise 1"
author: "patriciatrisno"
---

Getting Started

# Install and launching R packages

\^\*The code chunk below uses p_load() of pacman package to check if tidyverse packages have been in the computer. If they are, then they will be launced into R

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}

exam_data <- read_csv("~/Documents/SMU/April Term 2/Visual Analytics/patriciatrisno/ISSS608-VAA/Hands-on_Ex/Hands-on_Ex01/data/Exam_data.csv")
```

# Plotting a simple bar chart

```{r}
#| echo: false

ggplot(exam_data, aes(x = MATHS)) + 
  geom_bar()

ggplot(exam_data, aes(x = MATHS)) + 
  geom_dotplot(binwidth = 2.5, dotsize = 0.6) +
  scale_y_continuous(NULL, breaks = NULL)

ggplot(exam_data, aes(x = MATHS)) + 
  geom_histogram(bins = 20, color = "black", fill = "pink")

ggplot(exam_data, aes(x = MATHS, fill = GENDER)) + 
  geom_histogram(bins = 20, color = "grey30")

ggplot(exam_data, aes(x = GENDER, y = MATHS, fill = GENDER)) +
  geom_boxplot()


```

### **Geometric Objects**

```{r}
#| echo: false
ggplot(data=exam_data,
       aes(x = MATHS,
           colour = GENDER)) + 
  geom_density()

ggplot(data=exam_data,
       aes(y = MATHS,
           x = GENDER)) + 
  geom_boxplot(notch=TRUE)
```

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()

ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_point()

ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```

#### Working with ggplot2 elements - stat

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  stat_summary(geom = "point",
               fun = "mean",
               colour = "red",
               size=4) 

ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.2)

ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```

#### Working with ggplot2 elements - facet

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)

ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

#### Working with ggplot2 elements - coordinates

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()

ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)

ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

#### Working with ggplot2 elements - themes

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()

ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()

ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```
