---
title: "Take-home Exercise 1"
subtitle: "Beyond Basic ggplot2 Fundamentals"
description: "[ short describtion of the topic context ]"
author: patriciatrisno
date: "May 2, 2025"
date-modified: "last-modified"
format: html
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## 1 Overview

### 1.1 Background

--------------\[ describe the background context \]--------------

### 1.2 Understand the Data

--------------\[ explain the data sources \]--------------

From Singapore Department of Statistics

Title: **Singapore Residents by Planning Area/Subzone, Single Year of Age and Sex, June 2024**

Collected by with how many respondent

--------------\[ what is inside the data \]--------------

How many rows and column variables included below :

variables

----- \[ in table \]-----

PA -Planning Area -

SZ - Subzone -

Age - Single Year of Age -

Sex - Sex -

Pop - Resident Count -

Time - Time / Period -

--------------\[ data infos \]--------------

-   For June 2024, Planning Areas refer to areas demarcated in the Urban Redevelopment Authority's Master Plan 2019.

--------------\[ data limitation \]--------------

-   this data exclude residents who have been away from Singapore for a continuous period of 12 months or longer as at the reference period.

-   number had been rounded to nearest 10

### 1.3 Task

purpose and method

--------------\[ R packages used \]--------------

### 1.4 Working Plan and Content Directions

--------------\[ how each part contribute to the content creating \]--------------

--------------\[ Elaborate the directions \]--------------

## 2 Get Ready

::: panel-tabset
## Package Installations {.tabset}

\[ add code\]

If the package is already installed, p_load() will load it into your current R session. if it's not installed, p_load() will automatically download and install it from CRAN (the Comprehensive R Archive Network) before loading it.

```{r}
pacman::p_load(tidyverse)
```

## Package Description

{ list package used & its function \]

## Importing Data

\[ add code\]

```{r}

```
:::

## 3 Data Wrangling and Tidying

3.1 Checking Missing Value

3.2 Recoding Needed

## 4 Univariate Analysis

--------------\[ big picture on describing the use of vizualization to uncover..., understand..., etc \]--------------

4.1 Vizualiation 1

4.2 Vizualiation 2

4.3 Vizualiation 3

## 5 Exploratory Data Analysis

5.1 Vizualiation 1

5.2 Vizualiation 2

5.3 Vizualiation 3

## 6 Summary

--------------\[ add summary of findings \]--------------

## 7 Limitation and Future Study

--------------\[ limitation & future work \]--------------

-   only cover 2024 data and only until 24/9/2024

    which CANNOT represent the

-   As mentioned above also, this data exclude residents who have been away from Singapore for a continuous period of 12 months or longer as at the reference period.

    CANNOT

-   the number had been rounded to nearest 10

    Data CANNOT represent the true number of

## 8 Reference

--------------\[ references \]--------------
